Regression models for Cylindrical data in Psychology

### 2019-08-12 Manuscript/Manuscript.tex
Update manuscript + comments heleen

In the social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular component)
and a measure indicating the magnitude (i.e., the linear component). Many
established models in psychology are often referred to as circular or circumplex
models, but those models are cylindrical. Examples of such cylindrical models
are the interpersonal circle/circumplex
(Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans, \DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006), the circumplex of
affect (Russell, 1980), the circumplex of human emotion
(Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline
\indent Also, many of the more recent types of data that are studied in
psychology are cylindrical. For example, research on human navigation uses data
where distance (i.e., the linear component) and direction (i.e., the circular
component) are of interest (Chrastil \& Warren, 2017) or in eye-tracking, the
saccade data also consist of both the direction (i.e., the circular variable)
and the duration (i.e., the linear variable) (e.g., Rayner (2009)). Apart from
the social sciences, data with a circular and linear component more commonly
occur in meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \& Pereira, 2014) or marine
research (Lagona, Picone, Maruotti, \& Cosoli, 2015) \newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and
more of these researchers acknowledge that linear methods are not sufficient and
call for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill the
above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. These adapted cylindrical models are then used to analyse the
teacher data.

@@

\hfil [Figure \ref{circline} about here] \hfil

@@

first two moments, circular location or mean \(\mu\) and mean resultant length
\(\rho\)\footnote{In the von Mises distribution, a common distribution for a

@@

\hfil [Table \ref{TableModels} about here] \hfil

\subsubsection{The Modified Circular-Linear Projected Normal (CL-PN) and Modified Circular-Linear General Projected Normal (CL-GPN) Models}\label{CL-(G)PN}

@@

\hfil [Figure \ref{projection} about here] \hfil

@@

\hfil [Figure \ref{circregline} about here] \hfil

@@

\hfil [Figure \ref{plotGPN} about here] \hfil

@@

\subsubsection{The Modified Abe-Ley Model}\label{WeiSSVM}

@@

\section{Teacher Data}\label{Example}
\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Plots/IPC-T.png}
\caption{The interpersonal circle for teachers (IPC-T). The words presented in
the circumference of the circle are anchor words to describe the type of
behavior located in each part of the IPC.}
\label{QTI}
\end{figure}

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
\VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016) and Pennings (2017a). An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz and Strack (2011) for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) (Wubbels et al., 2006) which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. Agency refers
to the degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. This space is
called the interpersonal circle/circumplex (IPC) and different parts of this
space are characterized by different teacher behavior, e.g.~\enquote{helpful} or
\enquote{uncertain}. The IPC is ``a continuous order with no beginning or end''
(Gurtman, 2009, p. 2). We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable. This
circular variable represents location on the IPC of the interpersonal behavior
that a teacher shows towards his/her students.

\hfil [Figure \ref{QTI} about here] \hfil

\indent Cremers et al. (2018a) explain the circular nature of the IPC data and
analyze them as such using a circular regression model. The two dimension

@@

two-dimensional data are converted to the circle we lose some information,
namely the length of the two-dimensional vector \((A, C)^t\), \emph{i.e.}, its
Euclidean norm \(\mid\mid (A, C)^t \mid\mid\). This length represents the
intensity of the interpersonal behavior a teacher shows towards his/her
students. In a cylindrical model this intensity (the linear component) can be
modeled together with the location of interpersonal behavior of a teacher on the
IPC (the circular component). This leads to an improved analysis of interpersonal
circumplex data, over either analyzing the two dimensions separately or using a
circular model, because we take all information, circular and linear, into
account. In the next section we introduce several cylindrical models that can be
used to analyze the teacher data. First however we will provide descriptives for
the teacher data and conduct a \enquote{standard} analysis for these data that we can compare the results from the cylindrical models to.

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. This results in a sample of 151 teachers.
The data includes the location of interpersonal behavior on the IPC (IPC), the
circular component, and the intensity of interpersonal behavior (IPC intensity),
the linear component. It also includes teachers' self-efficacy (\verb|SE|)
concerning classroom management as covariate that is used to model the IPC and
IPC intensity. This means that \(\boldsymbol{x}_i = \boldsymbol{z}_i = (1,\: \text{SE}_i)\) in all cylindrical models except for the Abe-Ley model where
\(\boldsymbol{z}_i = \text{SE}_i\). In previous research, in psychology and
education it has been shown that higher self-efficacy is related to the quality
of interpersonal interactions (Locke \& Sadler, 2007; \VANDER{Want}{Van der}{van der} Want et al., 2018).
After listwise deletion of missings (\(3\) in total, only for the self-efficacy)
we have a sample of 148 teachers. Note that we remove the missings here for
simplicity, however, in general and especially when the number of missings is
larger, the influence of listwise deletion should be investigated and necessary
precaution (e.g.~imputation of missing values) should be taken. Table
\ref{Tableteacherdescriptives} shows descriptives for the dataset. Note that we
also show the scores on the Agency and Communion component of the IPC before
transformation to a circular score. For the circular variable IPC we show sample
estimates for the circular mean \(\bar{\theta}\) and mean resultant length
\(\hat{\rho}\). For the linear variables (Agency, Communion, IPC intensity and the
covariate SE) we show sample estimates of the linear mean and standard deviation
(sd). Figure \ref{dataplot} is a scatterplot showing the relation between the
linear and circular component of the teacher data for teachers with low SE
(below 1 sd below the mean), average SE (between 1 sd below and 1 sd above the
mean) and high SE (above 1 sd above the mean).

@@

\hfil [Table \ref{Tableteacherdescriptives} about here] \hfil

\hfil [Figure \ref{dataplot} about here] \hfil

@@

To perform a standard analysis of the teacher data we fit two linear regression
models to the Agency and Communion scores of the teachers separately. In these
regression models we incorporate an intercept and the covariate self-efficacy.
Coefficients and standard errors from the models are shown in Table
\ref{Tableteacherstandardanalysis}. From the results we conclude that the effect
of self-efficacy is significant for both the Agency and Communion component. An
increase of 1 unit on self-efficacy leads to a 0.07 increase in Agency and a
0.09 increase in Communion. Note that in this setup we can only model the effect
of self-efficacy on the intensity (size of the score) on the Agency and
Communion component. We can not quantify the effect on the location and
intensity on the IPC nor can we quantify their dependence.

\hfil [Table \ref{Tableteacherstandardanalysis} about here] \hfil

@@

\hfil [Table \ref{tab:estCLGPN}, \ref{tab:estAL} and \ref{tab:estCLGPNM} about here] \hfil

@@

\hfil [Figure \ref{reglineweib} about here] \hfil

@@

\hfil [Table \ref{tab:means} about here] \hfil

@@

\hfil [Table \ref{tab:ModelFit} about here] \hfil

@@

\hfil [Figure \ref{regline} about here] \hfil

@@

eye-tracking research. Furthermore, even though cylindrical models are already
used in fields outside of psychology, the addition of a regression structure to
the models is of use in these fields as well. \newline

@@

@@

@@

@@

\leavevmode\hypertarget{ref-wubbels2006interpersonal}{}%
Wubbels, T., Brekelmans, M., \DEN{Brok}{Den}{den} Brok, P., \& \VAN{Tartwijk}{Van}{van} Tartwijk, J. (2006). Handbook of classroom management: Research practice and contemporary issues. In C. Evertson \& C. S. Weinstein (Eds.), \emph{Handbook of classroom management: Research, practice, and contemporary issues} (pp. 1161--1191). Malwah, NJ: Lawrence Erlbaum Associates.

>>>>

In the social sciences the use of cylindrical data is very common. Cylindrical
data consist of a linear and a circular component. Gurtman (2011) refers
to cylindrical data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., the linear component).
Many established models in psychology are often referred to as circular or
circumplex models, but those models are cylindrical. Examples of such
cylindrical models are the interpersonal circle/circumplex
(Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans, \DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006), the circumplex of
affect (Russell, 1980), the circumplex of human emotion
(Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline
\indent Also, many of the more recent types of data that are studied in
psychology are cylindrical. For example, research on human navigation uses data
where distance (i.e., the linear component) and direction (i.e., the circular
component) are of interest (Chrastil \& Warren, 2017) or in eye-tracking, the
saccade data also consist of both the direction (i.e., the circular variable)
and the duration (i.e., the linear variable) (e.g., Rayner (2009)). Apart from
the social sciences, data with a circular and linear component more commonly
occur in meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \& Pereira, 2014) or marine
research (Lagona, Picone, Maruotti, \& Cosoli, 2015) \newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and
more of these researchers acknowledge that linear methods are not sufficient and
call for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill the
above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. These adapted cylindrical models are then used to analyse the
teacher data.

@@

\hfil \hspace{2cm} {[}Figure \ref{circline} about here{]} \hfil

@@

first two moments; the circular location or mean \(\mu\) (first moment) and mean resultant length
\(\rho\) (second moment)\footnote{In the von Mises distribution, a common distribution for a

@@

\hfil \hspace{2cm}{[}Table \ref{TableModels} about here{]} \hfil

\subsubsection{The modified Circular-Linear Projected Normal (CL-PN) and modified Circular-Linear General Projected Normal (CL-GPN) models}\label{CL-(G)PN}

@@

\hfil \hspace{2cm} {[}Figure \ref{projection} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Figure \ref{circregline} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Figure \ref{plotGPN} about here{]} \hfil

@@

\subsubsection{The modified Abe-Ley model}\label{WeiSSVM}

@@

\section{Teacher Data}\label{Example}
\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Plots/IPC-T2.png}
\caption{The interpersonal circle for teachers (IPC-T). The words presented in
the circumference of the circle are anchor words to describe the type of
behavior located in each part of the IPC.}
\label{QTI}
\end{figure}

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
\VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016) and Pennings (2017a). An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz and Strack (2011) for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) (Wubbels et al., 2006) which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. The item scores
on the QTI are measured on a scale from 1 to 5 and are later converted to a
scale ranging from -1 to 1. Subsequently they are converted to scores on Agency
and Communion (see also footnote 2 of Wubbels and Brekelmans (2005) for more details on this
procedure). Agency refers to the degree of power or control a teacher exerts in
interaction with his/her students. Communion refers to the degree of
friendliness or affiliation a teacher conveys in interaction with his/her
students. The loadings on the two dimensions of the QTI can be placed in a
two-dimensional space formed by Agency (vertical) and Communion (horizontal),
see Figure \ref{QTI}. This space is called the interpersonal circle/circumplex
(IPC) and different parts of the IPC are characterized by different teacher
behaviors, e.g.~\enquote{helpful} or \enquote{uncertain}. The IPC is ``a continuous order with no
beginning or end'' (Gurtman, 2009, p. 2). We call such ordering a
circumplex ordering and the IPC is therefore often called the interpersonal
circumplex. The circumplex ordering also implies that scores on the IPC could be viewed as
a circular variable. This circular variable represents location on the IPC of
the interpersonal behavior that a teacher shows towards his/her students.

\hfil \hspace{2cm} {[}Figure \ref{QTI} about here{]} \hfil

\indent Cremers et al. (2018a) and Cremers, Pennings, Mainhard, and Klugkist (2019) explain the circular nature of the IPC data and analyze them as such using a circular regression and mixed-effects model respectively. The two dimension

@@

two-dimensional data are converted to the location on the circle we lose some
information, namely the length of the two-dimensional vector \((A, C)^t\),
\emph{i.e.}, its Euclidean norm \(\mid\mid (A, C)^t \mid\mid\). This length
represents the intensity of the interpersonal behavior a teacher shows towards
his/her students. In a cylindrical model this intensity (the linear component)
can be modeled together with the location of interpersonal behavior of a teacher
on the IPC (the circular component). This leads to an improved analysis of
interpersonal circumplex data, over either analyzing the two dimensions
separately or using a circular model, because we take all information, circular
and linear, into account. In the next section we introduce several cylindrical
models that can be used to analyze the teacher data. First however we will
provide descriptives for the teacher data and conduct a \enquote{standard} analysis for
these data that we can compare the results from the cylindrical models to.

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. This results in a sample of 151 teachers.
The data includes the location of interpersonal behavior on the IPC (IPC), the
circular component, and the intensity of interpersonal behavior (IPC intensity),
the linear component. It also includes teachers' self-efficacy (\verb|SE|)
concerning classroom management as covariate that is used to model the IPC and
IPC intensity. This means that \(\boldsymbol{x}_i = \boldsymbol{z}_i = (1,\: \text{SE}_i)\) in all cylindrical models except for the Abe-Ley model where
\(\boldsymbol{z}_i = \text{SE}_i\). In previous research, in psychology and
education it has been shown that higher self-efficacy is related to the quality
of interpersonal interactions (Locke \& Sadler, 2007; \VANDER{Want}{Van der}{van der} Want et al., 2018).
After listwise deletion of missings (\(3\) in total, only for self-efficacy) we
have a sample of 148 teachers. We used listwise deletion of the missings in this
paper for simplicity. However, in general and especially when the number of
missings is larger, the influence of listwise deletion should be investigated
and necessary precaution (e.g.~imputation of missing values) should be taken.
Table \ref{Tableteacherdescriptives} shows descriptives for the dataset. Note
that we also show the scores on the Agency and Communion component of the IPC
before transformation to a circular score. For the circular variable IPC we show
sample estimates for the circular mean \(\bar{\theta}\) and mean resultant length
\(\hat{\rho}\). For the linear variables (Agency, Communion, IPC intensity and the
covariate SE) we show sample estimates of the linear mean and standard deviation
(sd). Figure \ref{dataplot} is a scatterplot showing the relation between the
linear and circular component of the teacher data for teachers with low SE
(below 1 sd below the mean), average SE (between 1 sd below and 1 sd above the
mean) and high SE (above 1 sd above the mean).

@@

\hfil \hspace{2cm} {[}Table \ref{Tableteacherdescriptives} about here{]} \hfil

\hfil \hspace{2cm} {[}Figure \ref{dataplot} about here{]} \hfil

@@

To perform a standard analysis of the teacher data we fit two linear regression
models to the Agency and Communion scores of the teachers separately. In these
regression models we incorporate an intercept and the covariate self-efficacy.
Coefficients and standard errors from the models are shown in Table
\ref{Tableteacherstandardanalysis}. From the results we conclude that the effect
of self-efficacy is significant for both the Agency and Communion component. An
increase of 1 unit on self-efficacy leads to a 0.07 increase in Agency and a
0.09 increase in Communion. Note that in this setup we can only model the effect
of self-efficacy on the intensity, the size of the score, of the Agency and
Communion component. We can not quantify the effect on the location and
intensity on the IPC nor can we quantify their dependence.

\hfil \hspace{2cm} {[}Table \ref{Tableteacherstandardanalysis} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Table \ref{tab:estCLGPN}, \ref{tab:estAL} and \ref{tab:estCLGPNM} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Figure \ref{reglineweib} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Table \ref{tab:means} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Table \ref{tab:ModelFit} about here{]} \hfil

@@

\hfil \hspace{2cm} {[}Figure \ref{regline} about here{]} \hfil

@@

eye-tracking research. For example, in a study by Pannasch, Helmert, Roth, Herbold, and Walter (2008), the relation between
saccade amplitudes and fixation duration is investigated in different
experimental settings. The saccade amplitude is a circular outcome that measures
the turning angle of the eye between fixations on different focal points, e.g.,
an individual first looks at one object in a picture and then moves focus to
another object. The outcome variables of interest are thus cylindrical in nature
(saccade amplitude is a circular outcome and fixation duration a linear one).
More importantly however, the researchers are interested in both the relation
between the circular and linear outcome and how covariates (different
experimental settings) affect the outcomes. These research questions can be
answered using the cylindrical models introduced in this paper. Furthermore,
even though cylindrical models are already used in fields outside of psychology,
the addition of a regression structure to the models is of use in these fields
as well. \newline

@@

\DeclareRobustCommand{\VANDER}[3]{#3} 
\DeclareRobustCommand{\VAN}[3]{#3}
\DeclareRobustCommand{\DEN}[3]{#3}

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
\setlength{\parskip}{8pt}

@@

\leavevmode\hypertarget{ref-cremers2019circular}{}%
Cremers, J., Pennings, H. J. M., Mainhard, T., \& Klugkist, I. (2019). Circular modelling of circumplex measurements for interpersonal behavior. \emph{Assessment}, OnlineFirst. doi:\href{https://doi.org/10.1177/1073191119858407}{10.1177/1073191119858407}

@@

\leavevmode\hypertarget{ref-pannasch2008duration}{}%
Pannasch, S., Helmert, J. R., Roth, K., Herbold, A.-K., \& Walter, H. (2008). Visual fixation durations and saccade amplitudes: Shifting relationship in a variety of conditions. \emph{Journal of Eye Movement Research}, \emph{2}(2). doi:\href{https://doi.org/10.16910/jemr.2.2.4}{10.16910/jemr.2.2.4}

@@

\leavevmode\hypertarget{ref-wubbels2005two}{}%
Wubbels, T., \& Brekelmans, M. (2005). Two decades of research on teacher--student relationships in class. \emph{International Journal of Educational Research}, \emph{43}(1), 6--24. doi:\href{https://doi.org/10.1016/j.ijer.2006.03.003}{10.1016/j.ijer.2006.03.003}

\leavevmode\hypertarget{ref-wubbels2006interpersonal}{}%
Wubbels, T., Brekelmans, M., \DEN{Brok}{Den}{den} Brok, P., \& \VAN{Tartwijk}{Van}{van} Tartwijk, J. (2006). An interpersonal perspective on classroom management in secondary classrooms in the Netherlands. In C. Evertson \& C. S. Weinstein (Eds.), \emph{Handbook of classroom management: Research, practice, and contemporary issues} (pp. 1161--1191). Malwah, NJ: Lawrence Erlbaum Associates.
